---
title: "Introduction to tidynhl"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to tidynhl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(width=1000)
set.seed(20201031L)
```

The goal of `tidynhl` is to make it easy for users to extract NHL data in a tidy format. In this sense, the word tidy refers to a state in which the data is
already neat enough to begin analyses right away, without having to clean anything or make any significant transformations.

This package was at first a project that I created to answer personal needs. This explains, in part, the dependency to the `data.table` package (I myself am
a massive user). Even if I could get rid of the dependency without significant efforts, I decided to stick to it, hoping to influence as many users as
possible to at least give it a try. This being said, it is totally possible to use the package without using `data.table` to handle the resulting data.
The native behavior of `tidynhl` is to return tidy datasets in `data.table` objects when the `data.table` package is attached to the current R session. In any
other cases, the returned datasets will be standard `data.frame` objects that could be handled directly with the native R functions, or transformed into
`tibble` objects for analysis with `dplyr`.

First of all, let's attach `tidynhl` and `data.table` to begin our overview of the package with a simple use case, analyzing the NHL careers of
[Vincent Lecavalier](https://www.nhl.com/player/vincent-lecavalier-8467329) and [Martin St. Louis](https://www.nhl.com/player/martin-st-louis-8466378), long
date teammates and important leaders of the [Tampa Bay Lightning](https://www.nhl.com/lightning) in the 2000s.

```{r setup}
library(tidynhl)
library(data.table)
```

The package contains two different families of functions:

- The functions prefixed by `tidy_` will return a tidy dataset ready to use in analyses.
- The functions prefixed by `get_` are helper functions to facilitate other tasks.

## NHL Entry Drafts

Even though Lecavalier and St. Louis both had successful careers, their NHL beginnings were totally opposite. Indeed, Lecavalier was the overall first pick of
the 1998 NHL Entry Draft while St. Louis was never drafted at all (even though he was available for selection in the 1993 NHL Entry Draft). First,
let's retrieve data about the 1993 and 1998 NHL Entry Drafts.

```{r retrieve_nhl_entry_drafts}
drafts <- tidy_draft(c(1993L, 1998L))
drafts[]
```

The top-5 selection the year when Lecavalier was drafted was

```{r lecavalier_top5}
drafts[draft_year==1998L & draft_overall<=5L]
```

We could make fun of NHL scouts by printing five random players that have been chosen in 1993 while St. Louis was still available.

```{r stlouis_random}
drafts[draft_year==1993L][sample(.N, 5L)]
```

I will admit that Brendan Morrison was a fair choice (still far behind St. Louis in terms of NHL success), but I honestly never heard of any other.

## Career overview

In order to be able to get information on Lecavalier and St. Louis, we will first need to get their personal NHL IDs.

```{r get_ids}
ids <- get_players_id(c("Vincent Lecavalier", "Martin St. Louis"))
ids
```

We can now retrieve some metadata about both players using those IDs.

```{r retrieve_meta}
meta <- tidy_players_meta(ids)
meta[]
```

Note the 8-inch difference in height! Let's now retrieve their NHL career statistics.

```{r retrieve_career_stats}
career_stats <- tidy_players_stats(ids)
career_stats[]
```

There is a lot of information above. Let's take a closer look at their total contributions as a Tampa Bay Lightning player in terms of some key statistics.

```{r tbl_stats}
career_stats[team_abbreviation=="TBL"][, .(
  nb_season = .N,
  games     = sum(skater_games),
  goals     = sum(skater_goals),
  assists   = sum(skater_assists),
  points    = sum(skater_points),
  plusminus = sum(skater_plusminus),
  pim       = sum(skater_pim)
), .(player_name, team_abbreviation, season_type)]
```

We could then conclude that, even though they were two excellent players for the Tampa Bay Lightning, St. Louis has a slight advantage over Lecavalier. The
biggest difference is probably their +/- ratio, which points out that St. Louis was probably a more responsible player in defense.

## 2004 Stanley Cup Playoffs

The Tampa Bay Lightning won their first Stanley Cup back in 2004. Lecavalier and St. Louis were key players in this conquest, but they both fell short of
winning the Conn Smythe Trophy, which has been awarded to [Brad Richards](https://www.nhl.com/player/brad-richards-8467389). So why not adding Richards in the
mix for the analysis of this subsection? No problem, let's do it! We begin by retrieving their game logs for this particular Stanley Cup Playoffs run.

```{r retrieve_gamelogs_stanley_cup}
stanleycup_gamelogs <- tidy_players_gamelogs(
  players_id = c(ids, get_players_id("Brad Richards")),
  seasons_id = "20032004",
  regular    = FALSE
)
stanleycup_gamelogs[]
```

Again, there is a lot of information above. Let's compare each player's success against their different opponents.

```{r compare_playoffs_opponent}
stanleycup_gamelogs[order(game_datetime)][, .(
  games     = sum(skater_games),
  goals     = sum(skater_goals),
  assists   = sum(skater_assists),
  points    = sum(skater_points),
  plusminus = sum(skater_plusminus),
  toi       = sum(skater_toi)
), .(player_name, opponent_abbreviation)]
```

Even though there is no objective way of ranking each player's contribution, one could argue that Lecavalier was the best player against the Montreal
Canadiens (2nd round), that St. Louis and Richards were kind of nose-to-nose against the New York Islanders (1st round) and the Philadelphia Flyers (Eastern
Semi-Final) and that Richards probably locked in the Conn Smythe by being the best player against the Calgary Flames in the Final.

Instead of splitting the data by opponents, we could have split it between home and away games to see if one was better than the others when playing at the
St. Pete Times Forum during this playoff year.

```{r compare_playoffs_homeaway}
stanleycup_gamelogs[, .(
  games           = sum(skater_games),
  goals           = sum(skater_goals),
  assists         = sum(skater_assists),
  points          = sum(skater_points),
  plusminus       = sum(skater_plusminus),
  points_per_game = sum(skater_points)/sum(skater_games)
), .(player_name, team_status)]
```

One could argue that, while Lecavalier was slightly better at home and St. Louis was slightly better away, Richards was clearly a more dominant player for that
playoff year while he was anywhere but in the Sunshine State.

## Conclusion

As we saw in this short introduction to `tidynhl`, the main goal of the package is to make it easy to retrieve NHL data, even for R beginning
users.
